Basic concepts and tools for the Toki Pona minimalist language: Wordnet synsets; analysis, synthesis and syntax highlighting of texts

### 2017-12-01 article/article.tex
encaminhada tabela de frequencia de letras

\section{Overview of the language}\label{basics}
This section describes very succinctly the formation of
words and sentences in the Toki Pona language.
It should enable a newcomer to grasp the essentials
of the language and the experienced to
acquire new insights.
Furthermore, it is a solid reference of the syntax and phonological
rules.

\subsection{Phonology}
Words in Toki Pona are written using only 14 letters:
\begin{itemize}
  \item Vowels a (open), e (mid front), o (mid back), i (close front),
    u (close back).
  \item Consonants j, k, l, m, n, p, s, t, w:
    \begin{itemize}
      \item Nasal: m (labial), n (coronal).
      \item Plosive: k (dorsal), p (labial), t (coronal).
      \item Fricative: s (coronal).
      \item Approximant: j (dorsal), l (coronal), w (labial).
    \end{itemize}
\end{itemize}

There are standard guidelines for pronunciation,
but the language allows for considerable allophonic
variation.
For example, /p t k s l/ might be pronounced
[p t k s l] or [b d g z \textipa{\!R}].
Especially for poetry, one might consider
j and w to be compound vowels
(e.g. j is 'i' and w is 'u'),
or as consonants.

@@

Vowel frequencies are as shown in Table~\ref{tab:vowels}
for starting, end and internal position.
Consonants are shown in Table~\ref{tab:cons}
for meaningful values.
Comparison of vowels agains consonant is in Table~\ref{tab:vcon}.
The high incidences of 'l' and 'a' favor
the naturalness of the language prosody
as it resembles lallation.

@@

\subsection{Syntax highlighting}
The same package~\cite{tokipona}
has a Vim syntax highlighting plugin
for Toki Pona.
Instructions for installing and using
the syntax highlighting is at~\cite{tokipona}.

@@

\section{Conclusions and further work}\label{conc}
\begin{itemize}
  \item Relate Toki Pona to Wordnet: should one Toki Pona word
    be related to more then one synset of the English language?
    An immediate implementation would be linking each Toki Pona word
    to the English words in the official vocabulary.
    A Toki Pona synset would be related to all the synsets of each English
    words regarded as lemmas (sequence of characters or sound).
  \item Translators. This Toki Pona Wordnet would be useful e.g. for naive translators
    of words and text.
    Enhancements of such translators would use POS tagging and templates
    in English and Toki Pona and/or corpus and further statistical methods.
    An efficiency of Toki Pona can be studied in terms of the coverage of
    the Wordnet cloud.
  \item Conlanging. In fact, I plan to use Wordnet for making new conlangs
    (conlanging), and insights derived from Toki Pona might be used for that end.
  \item Understand how the corpus is gathered in \url{tokipona.net}.
  \item Know about previously existing words that were used for Toki Pona
    (e.g. suno and suwi might come from sun and sweet),
    and about the reasons that lead Sonja (and maybe other people)
    to choose the 14 letters and the syllable structure.
    This might require a dedicated communication with the
    speaker community and the documentation authors.
  \item Corpus-based analysis.
  \item Publication of original texts and translations.
  \item Make an article written in Toki Pona.
    I wrote Section~\ref{finalToki}, and believe that
    a summary in both English and Toki Pona for
    facilitating the acquisition of context,
    a reasonable article on some scientific topics is
    possible.
    I first conceived something around complexity, statistics, physics, or computer science.
    But one possibility is to write about linguistics, philosophy, literature
    or psychology with the partners I write in English.
    I can start a draft,
    qhey might learn the language in a few hours (with or without my help),
    to contribute, and we can write a short paper.
  \item Enhance the synthesis of text to yield better contextualized text
    and stylistic traces, such as for poems and short stories.
    Give the user the ability to choose the sentences (generate randomly according to
    previously written or given text,
    some rules input by the user, the package and the language guidelines,
    outputs to the screen and asks to keep and discard).
  \item Enhance the syntax highlighting implementation described in Section~\ref{high}.
    It uses only syntactic cues to choose POS tags,
    which might be overcome by using n-grams and
    further techniques from Natural Language Processing~\cite{POS}.
  \item We are particularly interested in software tools,
    and one possibility is to use a conlang to describe software
    routines.
    In this sense, it seems profitable
    to understand what `non-ambiguous' means in Lojban,
    and in what sense one is able to compile and parse Lojban.
    This study will probably yield considerations about
    parsing Toki Pona files.
  \item Conlang idealized for describing data, programs,
    scientific writings, creative writing
    (exploring the thinking process).
    Different modes that might be specified by a section header or tag.
\end{itemize}

>>>>

\section{Overview of the language}\label{basics}
This section describes very succinctly the formation of
words and sentences in the Toki Pona language.
It should enable a newcomer to grasp the essentials
of the language and the experienced to
acquire new insights.
Furthermore, it is a solid reference of the phonological and syntax
rules.

\subsection{Phonology}
Words in Toki Pona are written using only 14 letters:
\begin{itemize}
  \item Vowels a (open), e (mid front), o (mid back), i (close front),
    u (close back).
  \item Consonants j, k, l, m, n, p, s, t, w:
    \begin{itemize}
      \item Nasal: m (labial), n (coronal).
      \item Plosive: p (labial), t (coronal), k (dorsal).
      \item Fricative: s (coronal).
      \item Approximant: w (labial), l (coronal), j (dorsal).
    \end{itemize}
\end{itemize}

There are standard guidelines for pronunciation,
but the language allows for considerable allophonic
variation.
For example, /p t k s l/ might be pronounced
[p t k s l] or [b d g z \textipa{\!R}].
Especially for poetry, one might consider
j and w to be vowels
(e.g. j as 'i' and w as 'u').

@@

Vowel frequencies are as shown in Table~\ref{freqLet}
for starting, end and internal position.
Consonants are shown in Table~\ref{tab:cons}
for meaningful values.
Comparison of vowels agains consonant is in Table~\ref{tab:vcon}.
The high incidences of 'l' and 'a' favor
the naturalness of the language prosody
as it resembles lallation.

@@

\subsection{Syntax highlighting}\label{shigh}
The same package~\cite{tokipona}
has a Vim syntax highlighting plugin
for Toki Pona.
Instructions for installing and using
the syntax highlighting is at~\cite{tokipona}.

@@

\section{Conclusions and further work}\label{conc}
\begin{itemize}
  \item Make a syntax highlighting that colors the tokens with respect
    to the syntatic position and function, and not in a fixed manner
    as is implemented and exposed in Section~\ref{shigh}.
  \item Relate Toki Pona to Wordnet: should one Toki Pona word
    be related to more then one synset of the English language?
    An immediate implementation would be linking each Toki Pona word
    to the English words in the official vocabulary.
    A Toki Pona synset would be related to all the synsets of each English
    words regarded as lemmas (sequence of characters or sound).
  \item Translators. This Toki Pona Wordnet would be useful e.g. for naive translators
    of words and text.
    Enhancements of such translators would use POS tagging and templates
    in English and Toki Pona and/or corpus and further statistical methods.
    An efficiency of Toki Pona can be studied in terms of the coverage of
    the Wordnet cloud.
  \item Conlanging. In fact, I plan to use Wordnet for making new conlangs
    (conlanging), and insights derived from Toki Pona might be used for that end.
  \item Understand how the corpus is gathered in \url{tokipona.net}.
  \item Know about previously existing words that were used for Toki Pona
    (e.g. suno and suwi might come from sun and sweet),
    and about the reasons that lead Sonja (and maybe other people)
    to choose the 14 letters and the syllable structure.
    This might require a dedicated communication with the
    speaker community and the documentation authors.
  \item Corpus-based analysis.
  \item Publication of original texts and translations.
  \item Make an article written in Toki Pona.
    I wrote Section~\ref{finalToki}, and believe that
    a summary in both English and Toki Pona for
    facilitating the acquisition of context,
    a reasonable article on some scientific topics is
    possible.
    I first conceived something around complexity, statistics, physics, or computer science.
    But one possibility is to write about linguistics, philosophy, literature
    or psychology with the partners I write in English.
    I can start a draft,
    qhey might learn the language in a few hours (with or without my help),
    to contribute, and we can write a short paper.
  \item Enhance the synthesis of text to yield better contextualized text
    and stylistic traces, such as for poems and short stories.
    Give the user the ability to choose the sentences (generate randomly according to
    previously written or given text,
    some rules input by the user, the package and the language guidelines,
    outputs to the screen and asks to keep and discard).
  \item Enhance the syntax highlighting implementation described in Section~\ref{high}.
    It uses only syntactic cues to choose POS tags,
    which might be overcome by using n-grams and
    further techniques from Natural Language Processing~\cite{POS}.
  \item We are particularly interested in software tools,
    and one possibility is to use a conlang to describe software
    routines.
    In this sense, it seems profitable
    to understand what `non-ambiguous' means in Lojban,
    and in what sense one is able to compile and parse Lojban.
    This study will probably yield considerations about
    parsing Toki Pona files.
  \item Conlang idealized for describing data, programs,
    scientific writings, creative writing
    (exploring the thinking process).
    Different modes that might be specified by a section header or tag.
\end{itemize}
